---
title: "NRE538_Data Munging"
author: "Oscar Feng-Hsun Chang"
date: "Week6"
output: 
  html_document:
    code_folding: show
    highlight: textmate
    keep_md: yes
    number_sections: true
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
      toc_depth: 4
---

We have seen some basic functions built in the _base_ package of __R__ (which means you don't have to install anything). However, there are several packages that allows you to clean/reorganize/munge the data in a more elegant and efficient fashion. They are [__reshape2__](https://cran.r-project.org/web/packages/reshape2/reshape2.pdf), [__plyr__](https://cran.r-project.org/web/packages/plyr/plyr.pdf),
[__dplyr__](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html), or 
[__tidyr__](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf) packages. There even exists a cheat sheet for [__dplyr__ and __tidyr__](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) packages.

We've seen `order`, `subset`, `r/cbind`, and `merge` before. Let's explore some more handy functions in the __reshape2__ and __plyr__ packages.

```{r}
library(reshape2)
library(plyr)
library(magrittr)
```

```{r, library loading, echo=FALSE, message=FALSE, warning=FALSE}
library(rmarkdown)
library(knitr) # Run R Code Chunks
library(ggplot2) # For plotting
library(RCurl)
```


```{r}
RY=read.csv(text=getURL("https://raw.githubusercontent.com/OscarFHC/NRE538_GSRA/master/Labs/NRE538_DataWrangling/RY.csv"), sep=",", header=T,comment.char="#")

id=read.csv(text=getURL("https://raw.githubusercontent.com/OscarFHC/NRE538_GSRA/master/Labs/NRE538_DataWrangling/sp_id.csv"), sep=",", header=T,comment.char="#")

trait=read.csv(text=getURL("https://raw.githubusercontent.com/OscarFHC/NRE538_GSRA/master/Labs/NRE538_DataWrangling/sp_stoi_traits.csv"), sep=",", header=T,comment.char="#")
```

```{r, echo=FALSE}
goal=read.csv(text=getURL("https://raw.githubusercontent.com/OscarFHC/NRE538_GSRA/master/Labs/NRE538_DataWrangling/goal.csv"), sep=",", header=T,comment.char="#")
```

Our goal is to organize the three data frames into the `goal` data frame.  
Following are some functions for organizing/cleaning/munging data in R. 

```{r}
head(goal)
```

# mutate

Given a data frame, `mutate()` modifies, adds, or removes variables.

```{r}
head(RY)

RY = RY %>%
  mutate(avgry = (RY1+RY2)/2)
head(RY)

RY = RY %>%
  mutate(avgry = avgry/2)
head(RY)
```

# subset

This allows you to `subset()` rows or `select` (this is an argument in `subset()`, NOT a function here!) columns.

```{r}
sp1 = RY %>%
  subset(sp1==1, select=c(RY1))
head(sp1)
```

This can also be achieved by the `filter()` (for rows) and `select()` (for columns) functions in the __dplyr__ package.

# merge/join

These two, by its name, do the *join* operation.
The `merge()` belongs to the __base__ package, and the `join()` belongs to __plyr__.

```{r}
RY.m = RY %>%
  merge(id, by.x="sp2", by.y="id") %>%
  merge(id, by.x="sp1", by.y="id") %>%
  arrange(sp1)

colnames(RY.m)[c(7:8)]=c("sp2_name", "sp1_name")
head(RY.m)
```

In the `join()`, we can specify the `type`, which is the same with the `all` argument in the `merge()` function. Try `? join` to understand it. 

```{r}
colnames(trait)[1] = "sp"
trait.1 = trait %>%
  join(id, by="sp", type="left")
  # merge(id, by="sp", all.x=TRUE) This and the above arguments will do the same thing.
head(trait.1)
```

Let's merge the two data frames but exclude species that have no trait information. We should have the following data frame. 

```{r, echo=FALSE}
trait.2 = trait %>%
  join(id, by="sp", type="inner")
  # merge(id, by="sp", all=FALSE) This and the above arguments will do the same thing.
  # subset(select=c(id, C.N, C.P, N.P))
head(trait.2)
```

# ddply

This is a very useful function that allows you to apply a function to a data frame and reture a data frame back.  
I always use this to summarize my own data. 

```{r}
sum = RY.m %>%
  ddply(~sp1, summarize,
        mean=mean(RY1),
        med=quantile(RY1, 0.5),
        lo=quantile(RY1, 0.05),
        hi=quantile(RY1, 0.95))
sum
```

# melt/cast

`melt` function can make a data frame from "wide" form into a "long" form.

```{r}
RY.long = RY %>%
  melt(id.vars=c("sp1", "sp2"), measure.vars=c("RY1", "RY2"), variable.name="RY_type", value.name="value")
head(RY.long)
```

`cast` function can do the reverse of `melt`

```{r}
RY.wide = RY.long %>%
  dcast(sp2+sp1~RY_type)
head(RY.wide)
```

# The GOAL

1. Your turn. Use what we've learned to generate the `goal` data frame. 

```{r, echo=FALSE}
final = RY.m %>%
  subset(select=c(1:5)) %>%
  merge(trait.2, by.x="sp2", by.y="id") %>%
  merge(trait.2, by.x="sp1", by.y="id") %>%
  arrange(sp1)
colnames(final)[6:13] = c("sp2.name", "C.N.sp2", "C.P.sp2", "N.P.sp2", "sp1.name", "C.N.sp1", "C.P.sp1", "N.P.sp1")
head(final)
```

In addition, if you are dealing with large dataset, [__data.table__](https://cran.r-project.org/web/packages/data.table/index.html) package will be your good friend!

2. Conduct some analysis to answer the following questions:   
2.1 Does different species have different relative yield (RY)? _hint_: ANOVA  
2.2 Does the RY of ones species depend on the other species? _hint_: interaction term  
2.3 Does different species have different C:N, C:P, N:P ratio? _hint_: ANOVA or MANOVA
```{r, include=FALSE}
RYaov = lm(RY1~as.factor(sp1)*as.factor(sp2), data=final)
summary.aov(RYaov)
```  
