---
title: "Week 4"
output: html_document
---

```{r setup, eval=FALSE}
setwd("~/Nat Resource Stats R")
getwd
```

```{r}
library(RCurl)
```


```{r, message=FALSE}
library(RCurl)
guns=read.csv(text=getURL("https://raw.githubusercontent.com/OscarFHC/NRE538_GSRA/master/Labs/NRE538_Lab2/guns.csv"), sep=",", header=T, comment.char="#")

```

One Sample t-test
```{r}
guns.MI = subset(guns, state=="Michigan")
t.test(guns.MI[,"murder"], mu=9)
```

Exercise 1
Null Hypotheses One Tail:
The mean robery rate in Michigan state over the years is not greater than 100. 

```{r}

t.test(guns.MI[,"robbery"], mu=100, alternative = 'greater')
```
Data from t.test:
data:  guns.MI[, "robbery"]
t = 13.768, df = 22, p-value = 1.361e-12
alternative hypothesis: true mean is greater than 100
95 percent confidence interval:
 217.2223      Inf
sample estimates:
mean of x 
 233.9261 
 
From this data, we recieved a p-value of 1.361e-12, which is well under the p-value of 0.05. With the p-value we recieved, we can safely reject the null hypothesis. This is because the p-value is below 0.05, and we can infact infer that the robbery rate in Michigan state is greater than 100. 


Null Hypotheses Two Tail: 
The mean robery rate over the years is equal to 100. 
```{r}
t.test(guns.MI[,"robbery"], mu=100, alternative='two.sided')
```
data:  guns.MI[, "robbery"]
t = 13.768, df = 22, p-value = 2.721e-12
alternative hypothesis: true mean is not equal to 100
95 percent confidence interval:
 213.7522 254.1000
sample estimates:
mean of x 
 233.9261 
 
After running the two-tailed t.test, we have recieved a p-value of 2.721e-12, which is less than 0.05. This means that we can reject our null hypothesis, our mean will not be equal to 100. We can safely then conclude that the mean robbery rate in the state of Michigan is not equal to 100. 


1.2 Two Sample t-test:
```{r}
guns.WAS = subset(guns, state=="Washington")
guns.MIN = subset(guns, state=="Minnesota")
t.test(guns.WAS[,"murder"], guns.MIN[,"murder"], paired=FALSE)


hist(guns.WAS[,"murder"], breaks=15, col="light blue", xlim=c(0, 7), ylim=c(0,5), xlab="murder rate", main="")
abline(v=mean(guns.WAS[,"murder"]), col="blue")
par(new=TRUE)
hist(guns.MIN[,"murder"], breaks=15, col="light green", xlim=c(0, 7), ylim=c(0,5), xlab="", main="")
abline(v=mean(guns.MIN[,"murder"]), col="dark green")
  legend("topright", legend = c("Washington mean", "Minnesota mean"),text.col=c("blue", "dark green"))
  
shapiro.test(guns.WAS[,"murder"])
shapiro.test(guns.MIN[,"murder"])
var.test(guns.WAS[,"murder"], guns.MIN[,"murder"])
```


1.3.1 One Sample Comparison
```{r}
muA=4
sdA=1
A=rnorm(30, mean=muA, sd=sdA)
A.mean=mean(A)

A.mean.SE=sqrt(var(A)/length(A))
A.mean.CI=c(A.mean-qnorm(0.975)*A.mean.SE, A.mean+qnorm(0.975)*A.mean.SE)

xs = seq(0, 15, by =0.01)
fA = dnorm(xs, mean=A.mean, sd=A.mean.SE)
plot(fA~xs, type="l", lwd=3, col="blue", xlim=c(3.5, 5), ylim=c(0, 3), 
     main="distribution of means of sample A", xlab="value", ylab="density")
abline(v=c(qnorm(0.025, mean=A.mean, sd=A.mean.SE), qnorm(0.975, mean=A.mean, sd=A.mean.SE)), col="blue", lty="dashed")
abline(v=4.5, col="dark green", lty="solid", lwd=3)

p.m.=2*(1-pnorm(4.5, mean=A.mean, sd=A.mean.SE))
p.m.

t.test(A, mu=4.5)$p.value
```

Exercise 2:

> You should use either resampling method or normal distribution assumption to claculate the p-value manually, i.e. not using the t.test nor the formula of t-test. 

```{r}
t.test(guns.MIN[,"murder"],guns.WAS[,"murder"], paired=FALSE, var.equal = TRUE)
```

Two Sample t-test

data:  guns.MIN[, "murder"] and guns.WAS[, "murder"]
t = -12.104, df = 44, p-value = 1.347e-15
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -2.454717 -1.753979
sample estimates:
mean of x mean of y 
 2.669565  4.773913 
 
From this test we see that the p-value is 1.347e-15. This means that we can reject the null hypotheses in that the murder rates of Washington and Minnesota are the same. This is because the p-value is less than 0.05, and so we reject the null hypotheses. We can therefore conclude that the murder rates in Washington state and the murder rates in Minnesota state are significantly different from each other. 