---
title: "Selva_2"
output: html_document
---

```{r}
library(Lahman)
data("Salaries")
money15 = subset(Salaries, yearID==2015)
head(money15)
```

```{r}
hist(money15$salary, main = "distribution of salary")
avg=mean(money15$salary)
  SD=sd(money15$salary)
  abline(v=avg, col="blue")
  legend("topright", legend = c(paste0("mean=", avg), paste0("SD=", SD)),text.col=c("blue", "dark green"))
```

Exercise 1:

```{r}
money14 = subset(Salaries, yearID==2014)
head(money14)
```

```{r}
meansVector = function(times, size, dat, varb){
  a = as.numeric(times)
  b = as.numeric(size)
  v = c()
  for(i in 1:a){
    y = sample(dat[,varb],b,replace=TRUE)
    m = mean(y)
    v = c(v,m)
  }
  v
}
```

Fixed number of values in each subsample, with increasing number of subsamples:
```{r}
means = meansVector(20, 100, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(40, 100, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(100, 100, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(500, 100, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(1000, 100, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(2000, 100, money14, "salary")
hist(means)  
```

Fixed number of subsamples, with increasing number of values in each subsample:

```{r}
means = meansVector(2000, 10, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(2000, 20, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(2000, 50, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(2000, 100, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(2000, 500, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(2000, 1000, money14, "salary")
hist(means)  
```

```{r}
means = meansVector(2000, 2000, money14, "salary")
hist(means)  
```

END EXERCISE 1


Exercise 2:

```{r}
data(mtcars)
```

```{r}
hist(mtcars$mpg, main = "distribution of mpg", breaks=10)
  avg=mean(mtcars$mpg)
  SD=sd(mtcars$mpg)
  abline(v=avg, col="blue")
  legend("topright", legend = c(paste0("mean=", avg), paste0("SD=", SD)),text.col=c("blue", "dark green"))
```

```{r}
meansVector = function(times, size, dat, varb){
  a = as.numeric(times)
  b = as.numeric(size)
  v = c()
  for(i in 1:a){
    y = sample(dat[,varb],b,replace=TRUE)
    m = mean(y)
    v = c(v,m)
  }
  v
}
```

Fixed number of values in each subsample, with increasing number of subsamples:
```{r}
means = meansVector(20, 100, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(40, 100, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(100, 100, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(200, 100, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(500, 100, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(1000, 100, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 100, mtcars, "mpg")
hist(means)  
```

Fixed number of subsamples, with increasing number of values in each subsample:
```{r}
means = meansVector(2000, 10, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 20, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 50, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 100, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 200, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 500, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 1000, mtcars, "mpg")
hist(means)  
```

```{r}
means = meansVector(2000, 2000, mtcars, "mpg")
hist(means)  
```

END EXERCISE 2

```{r}
library(Lahman)
data(Batting)
bat15 = subset(Batting, yearID==2015 & AB>200)
bat15$avg = bat15$H/bat15$AB
hist(bat15$avg, breaks=30, freq=FALSE)
lines(density(bat15$avg), col="red")
```

```{r}
shapiro.test(bat15$avg)
qqnorm(bat15$avg); qqline(bat15$avg, col="Red")
```

```{r}
avg.se = sd(bat15$avg)/sqrt(length(bat15$avg))
avg.se
ci.lo = mean(bat15$avg)-qnorm(0.95)*avg.se
ci.lo
ci.hi = mean(bat15$avg)+qnorm(0.95)*avg.se
ci.hi
```

```{r}
set.seed(61)
avg.m=c()
for (i in 1:10000){
  avg.m[i]=mean(sample(bat15$avg, length(bat15$avg), replace=TRUE))
}

mean.resamp=mean(avg.m)
mean.SE.resamp=sd(avg.m)
mean.CI.resamp=c(sort(avg.m)[10000*0.025], sort(avg.m)[10000*0.975])
print(paste0("standard error calculated by resampling = ", round(mean.SE.resamp,5)))
```

```{r}
print(paste0("95% confidence interval by resampling : ", 
             round(sort(avg.m)[10000*0.025],4), " - ", round(sort(avg.m)[10000*0.975],4)))
```

