---
title: "mmselva_7"
output: html_document
---
Exercise 4:

```{r}
mod4=lm(Temp~Ozone, data=airquality)
res4 = residuals(mod4)
RSS4=sum(res4^2)
k=2
n=length(res4)

mod.ex=subset(airquality, Ozone!= "NA")

SSY4=deviance(lm(Temp~1, data=mod.ex))
SSY4.1=var(mod.ex$Temp)*115

adjR2= 1 - (RSS4/(n-k))/(SSY4.1/(n-1))
adjR2
```

END EXERCISE 4

```{r}
data("airquality")
head(airquality, 15)

mod1=lm(Temp~Ozone, data=airquality)
summary(mod1)
plot(Temp~Ozone, data=airquality)
abline(lm(Temp~Ozone, data=airquality), col="red")
```

Exercise 5:

a.
```{r}
library(lmtest)
dwtest(mod1, alternative=c("two.sided"))

```
When running this test for autocorrelation, a signficant result will allow us to reject the null hypothesis that there is no auto correlation. We do not want autocorrelation. Therefore, the fact that we get a significant p-value of 7.276e-06 means that there is auto correlation, and we should manipulate the data before fiting a linear model.

b.
```{r}
plot(residuals(mod1)~fitted(mod1))
abline(lm(residuals(mod1)~fitted(mod1)), col="red")
bptest(mod1)

```
This second test checks for homoschedasticity of the residuals, which essentially means that the residuals are predictably different sizes as you move along the x axis. Usually, they start out small and gradually get bigger, which we do not want. The p-value of this test is not signficant, so the residuals of this model do not display homoschedasticity, which is what we want.

c.
```{r}
qqnorm(residuals(mod1))
qqline(residuals(mod1), col="red")

shapiro.test(residuals(mod1))
```
These test check for normality of the residuals. The qqnorm visually shows us that the residuals are not quite normal because they don't follow the 1:1 line. The Shapiro-Wilk test further tells us with a significant p-value, that the residuals are not normally distributed.

END EXERCISE 5