---
title: "juliaent_7"
output: html_document
---

```{r}
library("dplyr")
data("airquality")
airquality = airquality %>%
  subset(is.na(Solar.R)==FALSE & is.na(Ozone)==FALSE)
```

## Exercise 1
Manipulate the above code to reduce the correlation between X1 and X2 and recalculate the regression coefficients. Explain what you find.

hint: Increase the random noise of X1 and X2 so that the two can be less correlated.
```{r}
Y = seq(from=0, to=20, by=0.1)
X1 = (Y-2)/3 + runif(length(Y), min=0.1, max=10)
X2 = (Y+3)/2 + runif(length(Y), min=0.1, max=10)
df = as.data.frame(cbind(Y, X1, X2))
summary(lm(Y~X1, data=df))
summary(lm(Y~X2, data=df))
summary(lm(Y~X1+X2, data=df))
plot(X1, X2)
cor(X1, X2)
```
## As the min decreases and max increases, X1 and X2 become less correlated, the plot becomes less linear and correlation coefficient gets smaller. 


## Exercise 2
[Bonus, 2 pts] Challenge yourself by calculateing the RSS of the two models, F statistics (1 pt) and the p-value (1 pt) all by yourself (i.e. not using the anova() function)

Compare a third model (mod3) with ozone varaible as the third independent variable to mod1 and mod2 above. Does it explain more variance? Does it give you reasonable estimate of regressino coefficients?
Important F-test can only be used to compare the ?€œnested?€? models!
```{r}
mod1 = lm(Temp~Wind, data=airquality)
summary(mod1)
mod2 = lm(Temp~Wind+Solar.R, data=airquality)
summary(mod2)
mod3 = lm(Temp~Wind+Solar.R+Ozone, data = airquality)
summary(mod3)
anova(mod1, mod2, mod3)
```
## From the output table, we see that the RSS of mod3 is less than that of mod1 and mod2. The interpretation of these results is that mod3 explains a significantly greater amount of varaince compared to mod1 and mod2, because the probability for mod1 or mod 2 to explain the same amount of variance as mod3 is shown by the p values. The F statistic is 42.463.


## Exercise 3

Compare the AIC of the three models above mod1, mod2, and mod3. Do you have different conclusion in terms of which model performs better?
```{r}
AIC(mod1, mod2, mod3)
```
## Mod3 is the best preforming model because it has the smallest AIC value
